---
title: "Michelson Speed-of-light Measurements"
author: "Swasti Jain"
date: 2025
output: 
  github_document:
    toc: true
prerequisites:
  - e-data02-derive
editor_options: 
  markdown: 
    wrap: 72
---

*Purpose*: When studying physical problems, there is an important
distinction between *error* and *uncertainty*. The primary purpose of
this challenge is to dip our toes into these factors by analyzing a real
dataset.

*Reading*: [Experimental Determination of the Velocity of
Light](https://play.google.com/books/reader?id=343nAAAAMAAJ&hl=en&pg=GBS.PA115)
(Optional)

<!-- include-rubric -->

# Grading Rubric

<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics
define how you will be graded, both on an individual and team basis.

## Individual

<!-- ------------------------- -->

| Category | Needs Improvement | Satisfactory |
|----|----|----|
| Effort | Some task **q**'s left unattempted | All task **q**'s attempted |
| Observed | Did not document observations, or observations incorrect | Documented correct observations based on analysis |
| Supported | Some observations not clearly supported by analysis | All observations clearly supported by analysis (table, graph, etc.) |
| Assessed | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support |
| Specified | Uses the phrase "more data are necessary" without clarification | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Submission

<!-- ------------------------- -->

Make sure to commit both the challenge report (`report.md` file) and
supporting files (`report_files/` folder) when you are done! Then submit
a link to Canvas. **Your Challenge submission is not complete without
all files uploaded to GitHub.**

```{r message=FALSE, warning=FALSE}
# Libraries
library(tidyverse)
library(googlesheets4)

url <- "https://docs.google.com/spreadsheets/d/1av_SXn4j0-4Rk0mQFik3LLr-uf0YdA06i3ugE6n-Zdo/edit?usp=sharing"

# Parameters
LIGHTSPEED_VACUUM <- 299792.458 # Exact speed of light in a vacuum (km / s)
LIGHTSPEED_MICHELSON <- 299944.00 # Michelson's speed estimate (km / s)
LIGHTSPEED_PM <- 51 # Michelson error estimate (km / s)
```

*Background*: In 1879 Albert Michelson led an experimental campaign to
measure the speed of light. His approach was a development upon the
method of Foucault[3], and resulted in a new estimate of
$v_0 = 299944 \pm 51$ kilometers per second (in a vacuum). This is very
close to the modern *exact* value of `r LIGHTSPEED_VACUUM`. In this
challenge, you will analyze Michelson's original data, and explotre some
of the factors associated with his experiment.

I've already copied Michelson's data from his 1880 publication; the code
chunk below will load these data from a public googlesheet.

*Aside*: The speed of light is *exact* (there is **zero error** in the
value `LIGHTSPEED_VACUUM`) because the meter is actually
[*defined*](https://en.wikipedia.org/wiki/Metre#Speed_of_light_definition)
in terms of the speed of light!

```{r read-sheet}
## Note: No need to edit this chunk!
gs4_deauth()
ss <- gs4_get(url)
df_michelson <-
  read_sheet(ss) %>%
  select(Date, Distinctness, Temp, Velocity) %>%
  mutate(Distinctness = as_factor(Distinctness))

df_michelson %>% glimpse()
```

*Data dictionary*:

-   `Date`: Date of measurement
-   `Distinctness`: Distinctness of measured images: 3 = good, 2 = fair,
    1 = poor
-   `Temp`: Ambient temperature (Fahrenheit)
-   `Velocity`: Measured speed of light (km / s)

### **q1** Re-create the following table (from Michelson (1880), pg. 139) using `df_michelson` and `dplyr`. Note that your values *will not* match those of Michelson *exactly*; why might this be?

| Distinctness | n   | MeanVelocity |
|--------------|-----|--------------|
| 3            | 46  | 299860       |
| 2            | 39  | 299860       |
| 1            | 15  | 299810       |

```{r}
## TODO: Compute summaries
df_q1 <- df_michelson %>%
  group_by(Distinctness) %>%
  summarize(
    n = n(),
    mean_v = mean(Velocity),
  )
df_q1 %>%
  arrange(desc(Distinctness)) %>%
  knitr::kable()
```

**Observations**:

\- Write your observations here! -

Overall the numbers seem pretty close.

-   Why might your table differ from Michelson's?

These numbers might be affected by error during Michelson's
experimentation and data collection or he rounded his numbers when
documenting them. Considering these are huge numbers he may have
determined that the last digit was not so significant.

The Velocity likely correlates to the distinctness of the image. As
mentioned earlier, the levels of distinctness measure the level of
clarity from 3= good, 2=fair, 1=poor. It makes sense that the higher
level of distinctness may correlate to a more accurate velocity reading.

The `Velocity` values in the dataset are the speed of light *in air*;
Michelson introduced a couple of adjustments to estimate the speed of
light in a vacuum. In total, he added $+92$ km/s to his mean estimate
for `VelocityVacuum` (from Michelson (1880), pg. 141). While the
following isn't fully rigorous ($+92$ km/s is based on the mean
temperature), we'll simply apply this correction to all the observations
in the dataset.

### **q2** Create a new variable `VelocityVacuum` with the $+92$ km/s adjustment to `Velocity`. Assign this new dataframe to `df_q2`.

```{r q2-task}
## TODO: Adjust the data, assign to df_q2
df_q2 <- df_michelson %>%
  mutate(VelocityVacuum = Velocity + 92)

df_q2
```

As part of his study, Michelson assessed the various potential sources
of error, and provided his best-guess for the error in his
speed-of-light estimate. These values are provided in
`LIGHTSPEED_MICHELSON`---his nominal estimate---and
`LIGHTSPEED_PM`---plus/minus bounds on his estimate. Put differently,
Michelson believed the true value of the speed-of-light probably lay
between `LIGHTSPEED_MICHELSON - LIGHTSPEED_PM` and
`LIGHTSPEED_MICHELSON + LIGHTSPEED_PM`.

Let's introduce some terminology:[2]

-   **Error** is the difference between a true value and an estimate of
    that value; for instance `LIGHTSPEED_VACUUM - LIGHTSPEED_MICHELSON`.
-   **Uncertainty** is an analyst's *assessment* of the error.

Since a "true" value is often not known in practice, one generally does
not know the error. The best they can do is quantify their degree of
uncertainty. We will learn some means of quantifying uncertainty in this
class, but for many real problems uncertainty includes some amount of
human judgment.[2]

### **q3** Compare Michelson's speed of light estimate against the modern speed of light value. Is Michelson's estimate of the error (his uncertainty) greater or less than the true error?

```{r q3-task}
## TODO: Compare Michelson's estimate and error against the true value
## Your code here!
df_q2 %>%
  summarize(
    error = LIGHTSPEED_VACUUM - LIGHTSPEED_MICHELSON,
    LIGHTSPEED_PM
  )
```

**Observations**:

\- Is Michelson's estimate of the error (his uncertainty) greater or
less than the true error?

Michelson's estimate of his error is far less than the true error.

\- Make a quantitative comparison between Michelson's uncertainty and
his error.

He predicted 51 km/s when that was 100 km/s away from the real range of
error. This is an immense error from his range of uncertainty. Nearly
three times the error he predicted. This could have extreme issues as
the error multiplies for more precise calculations.

The following plot shows all of Michelson's data as a [control
chart](https://en.wikipedia.org/wiki/Control_chart); this sort of plot
is common in manufacturing, where it is used to help determine if a
manufacturing process is under [statistical
control](https://en.wikipedia.org/wiki/Statistical_process_control).
Each dot is one of Michelson's measurements, and the grey line connects
the mean taken for each day. The same plot also shows simulated data
using a probability model. We'll get into statistics later in the
course; for now, let's focus on understanding what real and simulated
data tend to look like.

### **q4** Inspect the following plot with the `Real` Michelson data and `Simulated` data from a probability model. Document the similarities and differences between the data under *observe* below.

```{r q4-cf-real-simulated}
## Note: No need to edit this chunk!
## Calibrate simulated data
v_mean <-
  df_q2 %>%
  summarize(m = mean(VelocityVacuum)) %>%
  pull(m)
v_sd <-
  df_q2 %>%
  summarize(s = sd(VelocityVacuum)) %>%
  pull(s)

## Visualize
set.seed(101)
df_q2 %>%
  mutate(Simulated = rnorm(n(), mean = v_mean, sd = v_sd)) %>%
  rename(Real = VelocityVacuum) %>%
  pivot_longer(
    cols = c(Simulated, Real),
    names_to = "source",
    values_to = "velocity"
  ) %>%
  ggplot(aes(Date, velocity)) +
  geom_hline(
    yintercept = LIGHTSPEED_MICHELSON,
    linetype = "dotted"
  ) +
  geom_hline(
    yintercept = LIGHTSPEED_MICHELSON - LIGHTSPEED_PM,
    linetype = "dashed"
  ) +
  geom_hline(
    yintercept = LIGHTSPEED_MICHELSON + LIGHTSPEED_PM,
    linetype = "dashed"
  ) +
  geom_line(
    data = . %>%
      group_by(Date, source) %>%
      summarize(velocity_mean = mean(velocity)),
    mapping = aes(y = velocity_mean),
    color = "grey50"
  ) +
  geom_point(
    mapping = aes(y = velocity),
    size = 0.8
  ) +
  facet_grid(source ~ .) +
  theme_minimal() +
  labs(
    x = "Date of Measurement (1879)",
    y = "Velocity (in Vacuum)"
  )
```

**Observations**:

Similarities -

The bounds of error are quite comparable. Both have the mean line around
299,950 km/s and the upper bound around 300,000 km/s and the lower bound
around 299,900 km/s.

Differences -

The outliers are more dramatic in the real measured data rather than
simulated data. The furthest outlier is over \~200 km/s above from the
mean line for the real data and the furthest outlier is under \~200 km/s
beneath the mean line.

### **q5** You have access to a few other variables. Construct a **at least three** visualizations of `VelocityVacuum` against these other factors. Are there other patterns in the data that might help explain the difference between Michelson's estimate and `LIGHTSPEED_VACUUM`?

```{r}
#
v_mean <- df_q2 %>%
  summarize(m = mean(VelocityVacuum)) %>%
  pull(m)
v_sd <- df_q2 %>%
  summarize(s = sd(VelocityVacuum)) %>%
  pull(s)

## Visualize
set.seed(101)
df_q2 %>%
  # mutate(Simulated = rnorm(n(), mean = v_mean, sd = v_sd)) %>%
  rename(Real = VelocityVacuum) %>%
  pivot_longer(
    cols = c(Real),
    names_to = "source",
    values_to = "velocity"
  ) %>%
  ggplot(aes(Temp, velocity)) +
  geom_hline(
    yintercept = LIGHTSPEED_MICHELSON,
    linetype = "dotted"
  ) +
  geom_hline(
    yintercept = LIGHTSPEED_MICHELSON - LIGHTSPEED_PM,
    linetype = "dashed"
  ) +
  geom_hline(
    yintercept = LIGHTSPEED_MICHELSON + LIGHTSPEED_PM,
    linetype = "dashed"
  ) +
  geom_hline(
    yintercept = LIGHTSPEED_MICHELSON + LIGHTSPEED_PM,
    linetype = "dashed"
  ) +
  geom_line(
    data = . %>%
      group_by(Temp, source) %>%
      summarize(velocity_mean = mean(velocity)),
    mapping = aes(y = velocity_mean),
    color = "grey50"
  ) +
  geom_point(
    mapping = aes(y = velocity),
    size = 0.8
  ) +

  # facet_grid(source~.) +
  theme_minimal() +
  labs(
    x = "Temperature (F)",
    y = "Velocity (in Vacuum)"
  )
```

```{r}
#
v_mean <-
  df_q2 %>%
  summarize(m = mean(VelocityVacuum)) %>%
  pull(m)
v_sd <-
  df_q2 %>%
  summarize(s = sd(VelocityVacuum)) %>%
  pull(s)

## Visualize
set.seed(101)
df_q2 %>%
  # mutate(Simulated = rnorm(n(), mean = v_mean, sd = v_sd)) %>%
  rename(Real = VelocityVacuum) %>%
  pivot_longer(
    cols = c(Real),
    names_to = "source",
    values_to = "velocity"
  ) %>%
  ggplot(aes(Temp, velocity, color = Distinctness)) +
  geom_hline(
    yintercept = LIGHTSPEED_MICHELSON,
    linetype = "dotted"
  ) +
  geom_hline(
    yintercept = LIGHTSPEED_MICHELSON - LIGHTSPEED_PM,
    linetype = "dashed"
  ) +
  geom_hline(
    yintercept = LIGHTSPEED_MICHELSON + LIGHTSPEED_PM,
    linetype = "dashed"
  ) +
  geom_hline(
    yintercept = LIGHTSPEED_MICHELSON + LIGHTSPEED_PM,
    linetype = "dashed"
  ) +
  geom_line(
    data = . %>%
      group_by(Temp, source) %>%
      summarize(velocity_mean = mean(velocity)),
    mapping = aes(y = velocity_mean),
    color = "grey50"
  ) +
  geom_point(
    mapping = aes(y = velocity),
    size = 0.8
  ) +

  # facet_grid(source~.) +
  theme_minimal() +
  labs(
    x = "Temperature (F)",
    y = "Velocity (in Vacuum)"
  )
```

```{r}
#
v_mean <-
  df_q2 %>%
  summarize(m = mean(VelocityVacuum)) %>%
  pull(m)
v_sd <-
  df_q2 %>%
  summarize(s = sd(VelocityVacuum)) %>%
  pull(s)

## Visualize
set.seed(101)
df_q2 %>%
  # mutate(Simulated = rnorm(n(), mean = v_mean, sd = v_sd)) %>%
  rename(Real = VelocityVacuum) %>%
  pivot_longer(
    cols = c(Real),
    names_to = "source",
    values_to = "velocity"
  ) %>%
  ggplot(aes(Temp, velocity, color = Distinctness)) +
  geom_hline(
    yintercept = LIGHTSPEED_MICHELSON,
    linetype = "dotted"
  ) +
  geom_hline(
    yintercept = LIGHTSPEED_MICHELSON - LIGHTSPEED_PM,
    linetype = "dashed"
  ) +
  geom_hline(
    yintercept = LIGHTSPEED_MICHELSON + LIGHTSPEED_PM,
    linetype = "dashed"
  ) +
  geom_hline(
    yintercept = LIGHTSPEED_MICHELSON + LIGHTSPEED_PM,
    linetype = "dashed"
  ) +
  geom_hline(
    yintercept = LIGHTSPEED_VACUUM,
    linetype = "dashed",
    color = "firebrick"
  ) +
  geom_line(
    data = . %>%
      group_by(Temp, source) %>%
      summarize(velocity_mean = mean(velocity)),
    mapping = aes(y = velocity_mean),
    color = "grey50"
  ) +
  geom_point(
    mapping = aes(y = velocity),
    size = 0.8
  ) +

  # facet_grid(source~.) +
  theme_minimal() +
  labs(
    x = "Temperature (F)",
    y = "Velocity (in Vacuum)"
  )
```

```{r}
df_q2 %>%
  rename(Real = VelocityVacuum) %>%
  pivot_longer(
    cols = c(Real),
    names_to = "source",
    values_to = "velocity"
  ) %>%
  ggplot(aes(velocity)) +
  geom_bar()
```

```{r}
df_q2 %>%
  rename(Real = VelocityVacuum) %>%
  pivot_longer(
    cols = c(Real),
    names_to = "source",
    values_to = "velocity"
  ) %>%
  ggplot(aes(Date, velocity, color = Distinctness)) +
  geom_point()
```

**Observations**:

-   Graph 1: Removed Simulated Data and Plotted Velocity (in Vacuum)
    against Temperature (in Fahrenheit)
    -   I chose to focus on the Real Data to understand the frame of
        reference Michelson was working in.
    -   Since Temperature is traditionally an independent variable, I
        believed it would have been illuminating to see the relationship
        between velocity and temperature.
        -   According to this graph it is hard to map any relationship
            between temperature affecting the velocity measurements
-   Graph 2: Added Distinctness Coloring
    -   Another element I wanted to see is if plotting the distinctness
        values would tell a different story
        -   There is a larger cluster of level 1 (poor) points under 70
            F however there isn't anything definitive from this either
            as the level 2 (fair) and level 3 (good) points are
            scattered with no discernible clusters.
-   Graph 3: Added LIGHTSPEED_VACUUM Horizontal Line
    -   Lastly I added a horizontal line for the lightspeed_vacuum value
        to see how far away the entire data set is from the 'real' mean
        value. It is interesting to see that the discrepancy is so
        disparate that all but two data points are over that plotted
        value.

## Bibliography

-   [1] Michelson, [Experimental Determination of the Velocity of
    Light](https://play.google.com/books/reader?id=343nAAAAMAAJ&hl=en&pg=GBS.PA115)
    (1880) 
-   [2] Henrion and Fischhoff, [Assessing Uncertainty in Physical
    Constants](https://www.cmu.edu/epp/people/faculty/research/Fischoff-Henrion-Assessing%20uncertainty%20in%20physical%20constants.pdf)
    (1986) 
-   [3] BYU video about a [Fizeau-Foucault
    apparatus](https://www.youtube.com/watch?v=Ik5ORaaeaME), similar to
    what Michelson used.
